---
title: "Engineering Equations using the `engcalc` Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Engineering-Math}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, message=FALSE, warning=FALSE, error=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  results = 'asis'
)
```

Symbolic mathematics are excellent for both showing working such as in engineering as well as when teaching. `engcalc` uses the `Ryacas` package to implement the `YACAS` software in `R` for symbolic mathematics, and adds the functionality of being able to show the substitution of values for variables without simplification for equations involving integration, differentiation and matrix operations.

It does this by catching the unevaluated arguments as strings and then performs substitutions on the strings at various stages of the calculation so that the substituted, but not solved equations can be seen. `engcalc` also enables variable names that involve subscript, which are common in engineering practice such as $\theta_{(x,z)}$, $\sigma_{(1,2)}$ to be used in the `Ryacas` workflow.

Whilst `engcalc` does not interfere with the `Ryacas` commands, it has been optimised to work with integrals, derivatives and simple matrix operations i.e. those found in typical engineering practice.

First, lets load some packages:

```{r setup}
library(engcalc)
library(Ryacas)
library(magrittr)
```

The following examples demonstrate some of the key features:

## Example 1

Here is a stress tensor $\sigma$, which includes components with names in the form of: $\sigma_{x,y}$, written `\sigma_{x, y}`, which is conducive to Latex typesetting, but not practical for `R` variable names

```{r}
xyz <- c('x', 'y', 'z')
s <-
  sapply(xyz, function(x)
    paste('sigma_{', xyz, ',', x, '}', sep = ''))
```

We can render this as Latex using the `tex` command from `Ryacas`:

```{r}
s %>%
  ysym() %>%
  tex() %>%
  paste0('$$ \\sigma = ', ., '$$') %>%
  cat()
```

Next, we will create an elasticity moduli vector $E$ which dictates the relationsship between $\sigma$ andthe strain $\varepsilon$:

```{r}
E <- paste('E_{', xyz, '}', sep = '') 
E %>%
  ysym() %>%
  tex() %>%
  paste0('$$ E = ', ., '$$') %>%
  cat()
```

And now we will use the `sub_eq` function from `engcalc` to solve for the strains $\varepsilon$ Youngs Moduli $E$ that dictates the replationship between the stress and strain: $\varepsilon = \frac{\sigma}{E}$.

### The `sub_eq` Function

The `sub_eq` function takes the following inputs:

1. `eq` the string representation of the equation to solve. In this case: $\sigma / E$.
1. `...` input of `Ryacas::ysym` variables which feature in `eq` or in the variables that `eq` refers to. In this case `eq` references `sigma` and `E`, so we will input `sigma = s` and `E = E`. Note that the names do not have to be the same.
1. `vars` an optional list which holds named `ysym` objects and variables which feature in `eq` or in the variables that `eq` refers to. In this case we could create a list i.e.: `vars = list(sigma = s, E = E)`, which is identical to inputting these as named arguments into `...`. 

We will tackle this problem using both `...` adn `vars` to input the arguments:

### `sub_eq` Method 1: Using `...`

```{r, results = 'asis'}
# first transform the string matrices to symbolic
sigma <- ysym(s)
E <- ysym(E)

varepsilon_1 <- sub_eq('sigma / E', sigma = s, E = E)

y <- lapply(varepsilon_1[1:3], function(x) {x %>%
  paste0('$$ \\varepsilon_1 = ', ., '$$') %>%
  cat()
})
```

### `sub_eq` Method 2: Using `vars`

```{r, results = 'asis'}
varepsilon_2 <- sub_eq('sigma / E', vars = list(sigma = s, E = E))

y <- lapply(varepsilon_2[1:3], function(x) {x %>%
  paste0('$$ \\varepsilon_2 = ', ., '$$') %>%
  cat()
})
```

Note how the results are identical:

```{r}
identical(varepsilon_1, varepsilon_2)
```

### `sub_eq` Method 3: Using `vars` and `...`

It is possible to use both `vars` and `...` to pass variables to `sub_eq`. The results so far have been symbolic representations of $\varepsilon$, so we will now assign some values to each of the 12 variables (9 in $\sigma$ and 3 in $\varepsilon$) to show some simple substitution usign both `vars` and `...`^[values of `1:12` are used for ease of seeing the substitutions]:

First, create some values for the variables in $\sigma$ and $E$:

```{r vals}
vals <- 1:12
names(vals) <- c(s, paste('E_{', xyz, '}', sep = '') )
knitr::kable(data.frame(Variable = paste('$\\', names(vals), '$'), Value = vals))
vals <- as.list(vals)
```

Now we will substitute these values into the equation and solve, showing the steps:

```{r, results = 'asis'}
# first transform the string matrices to symbolic
varepsilon_3 <- sub_eq('sigma / E', sigma = s, E = E, vars = vals)

y <- lapply(varepsilon_3, function(x) {x %>%
  paste0('$$ \\varepsilon_3 = ', ., '$$') %>%
  cat()
})
```


## Example 2 - Beam stiffness matrix multiplication

We know that $F=K\bullet d$ so lets create a beam stiffness matrix $K$ and displacement vector $d$ and use these to calculate the force vector $F$:

```{r, results = 'asis'}
# a structural engineering beam stiffness matrix
KK <- matrix(c('12 * EI/L^3', '6 * EI/L^2', '-12 * EI/L^3', '6 * EI/L^2',
               '6 * EI/L^2', '4 * EI/L', '-6 * EI/L^2', '2 * EI/L',
               '-12 * EI/L^3', '-6 * EI/L^2', '12 * EI/L^3', '-6 * EI/L^2',
               '6 * EI/L^2', '2 * EI/L', '-6 * EI/L^2', '4 * EI/L'),
             nrow = 4) %>% 
  ysym()

KK %>%
  tex() %>%
  paste0('$$K = ', ., '$$') %>%
  cat()

# a structural engineering beam displacement vector
# with typical notations for conversion to latex
dd <- c('v_{y,1}', 'theta_{z,1}', 'v_{y,2}', 'theta_{z,2}') %>%
  ysym()

dd %>%
  tex() %>%
  paste0('$$d = ', ., '$$') %>%
  cat()
```

We want to solve this using know quantities which we will now specify:

```{r}
val <- 3:8
names(val) <- c('v_{y,1}', 'theta_{z,1}', 'v_{y,2}', 'theta_{z,2}', 'EI', 'L')
print(val)
```


First, we can just simplify the matrix using these values:

```{r}
results <- sub_eq('K', K = KK, vars = val %>% as.list())
y <- lapply(results, function(x) {
  x %>%
  paste0('$$ K = ', ., '$$') %>%
  cat()
})
```

We will subsititute these values into the equation $F = K\bullet d$ and show the working:

```{r}
results <-
  sub_eq('K * d', d = dd, K = KK,
  vars = val %>% as.list()
)

cat('$$F = K \\bullet d$$')

KKK <- sub_eq('K', K = KK, vars = val %>% as.list())
ddd <- sub_eq('d', d = dd, vars = val %>% as.list())
cat(paste0('$$F = ', 
           KKK[[1]], 
           ' \\bullet ', 
           ddd[[1]], '$$'))

cat(paste0('$$F = ', 
           KKK[[2]], 
           ' \\bullet ', 
           ddd[[2]], '$$'))

y <- lapply(results, function(x) {
  x %>%
  paste0('$$ F = ', ., '$$') %>%
  cat()
})
```

We may only want to substitute some of the variables as there may still be some unknowns:

```{r}
# partial solution
results <-
  sub_eq('K * d', d = dd, K = KK,
  vars = val[5:6] %>% as.list()
)

y <- lapply(results, function(x) {
  x %>%
  paste0('$$ F = ', ., '$$') %>%
  cat()
})
```

Another example:


```{r}
results <- sub_eq('K * ds', K = KK, ds = dd, vars = val %>% as.list())
y <- lapply(results, function(x) {
  x %>%
  paste0('$$ F = ', ., '$$') %>%
  cat()
})
```

## Example 3 - Integration of a beam stiffness matrix

It is possible to use any of the `Ryacas` functions and show the substituted expression.

```{r}
# A more complex case involving integration
# results <-
#   sub_eq('integrate(K, "xx", 1, 3)',
#    K = KK, xx = 'xx', vars = val %>% as.list()
# )
# 
# y <- lapply(results, function(x) {
#   x %>%
#   paste0('$$ F = ', ., '$$') %>%
#   cat()
# })
```

$$\left( \begin{array}{c} \sigma _\left( x, x\right)  E_\left( x\right)  + \sigma _\left( x, y\right)  E_\left( y\right)  + \sigma _\left( x, z\right)  E_\left( z\right)  \\ \sigma _\left( y, x\right)  E_\left( x\right)  + \sigma _\left( y, y\right)  E_\left( y\right)  + \sigma _\left( y, z\right)  E_\left( z\right)  \\ \sigma _\left( z, x\right)  E_\left( x\right)  + \sigma _\left( z, y\right)  E_\left( y\right)  + \sigma _\left( z, z\right)  E_\left( z\right)  \end{array} \right)$$
